Processing genome scale tabular data with wormtable

### 2013-08-04 docs/paper/paper.tex
Added Tabix discussion and seek performance section.

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} 
      Biological research generates large amounts of tabular data.
These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is a time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. 
\paragraph*{Results:} 
We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables.

@@

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a database in 
which the keys are the values from the columns in question concatenated
together, and the values are pointers to the original row. Indexes make
many operations much more efficient, since we can directly seek to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns.

@@

\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using a simple interface. The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read (similar to 
a select list in SQL). Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings.
Providing an index as an argument to the \texttt{cursor} function
results in the rows being visited in the order defined by the 
index. The arguments \texttt{start} and \texttt{stop} 
then stipulate the range of values we are interested in.
There are several other useful classes and methods, but 
the \texttt{cursor} method encapsulates the core functionality
of wormtable: efficient iteration over rows.

The interface is kept as
simple and similar to the Python standard library as possible,
to ensure that there is little difficulty in learning the API.
Similarly, although one can readily imagine defining a 
query language using these 
primitive operations, we have deliberately avoided doing so. Adding 
an SQL-like query language would greatly increase the complexity 
of wormtable, and would only serve to raise the barrier to entry.
Less experienced programmers 
do not need to learn a query language to process data 
in a very high performance manner.

\subsection*{Examples}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable, we
converted a large 
publicly available VCF file, produced as part of the
Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a standard workstation with a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).

@@

Another advantage of wormtable is the ability to create 
an index over arbitrary combinations of columns within a table.
Indexes make many operations much more efficient, as they provide 
a way to quickly find particular rows and avoid the 
heavy cost of a full table scan.
To illustrate this advantage, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m40s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

\section*{Discussion}
It is well known that parsing text files is a very inefficient process, 
and binary file formats are preferable for this reason. BCF,
for example, is a binary version of the VCF format discussed above which 
is designed to avoid just this problem. There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for several different file formats is surely 
unsustainable.

@@

The library supports efficient access to any data stored in 
wormtable format, but there is currently a conversion tool for
the VCF format only. The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
are there other formats we should chuck in here?].

Other probable directions for future development of wormtable 
are to incorporate data file compression; to develop a 
C library and bindings for other popular languages such as 
Perl and R; and to generally enrich the API with features useful 
to the community. It is difficult to predict the 
precise direction of future development since this depends 
on community involvement.  Wormtable is an open and 
collaborative project, actively seeking feedback and 
contributors.

@@

Wormtable is not intended to replace text files as the universal 
interchange format for biological data. 
It is intended to provide a persistent data structure 
that can be efficiently processed 
and searched. Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write simple code to process 
data in a high performance manner. 
Different file formats can be handled consistently 
in wormtable, reducing the need for third party libraries 
to parse complex files and simplifying the code required 
to process data.
Together, these advantages of increased performance and reduced 
code complexity can substantially increase a researcher's 
productivity and ability to explore their data.

%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}
\paragraph{Project name:} Wormtable.
\paragraph{Project home page:}\url{https://pypi.python.org/pypi/wormtable}
\paragraph{Operating system(s):} POSIX.
\paragraph{Programming language:} C, Python.
\paragraph{Other requirements:} Berkeley DB 4.8 or later.
\paragraph{License:} GNU GPLv3+.
\paragraph{Any restrictions to use by non-academics:} None.

>>>>

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:} 
Biological research generates large amounts of tabular data.
These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is a time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. 
\paragraph*{Results:} 
We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables. Files written in existing 
formats can be easily converted to wormtable format, and we 
provide conversion utilities for the VCF and GTF format.

@@

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a database in 
which the keys are the values from the columns in question concatenated
together, and the values are pointers to the original row. Indexes make
many operations much more efficient, since we can directly seek to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns. Wormtable also supports binned indexes, in which a range of 
values is mapped to a single index key. This is very useful for floating 
point columns, where it may not be necessary to distinguish between
values that differ by small amounts.

@@

\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which dictate 
rows of interest. This allows us to efficiently seek to an 
arbitary location in the table and to read a given number 
of rows sequentially from this point.

The \texttt{Index} class also has a \texttt{cursor} function
with the same signature, but in this case, rows 
are returned in the order defined by the index. The  
\texttt{start} and \texttt{stop} arguments are now defined 
in terms of index keys: all rows in which the index key
is greater than or equal to \texttt{start} and less 
than \texttt{stop} are returned. Partial keys may also be 
provided for multi-column indexes. The \texttt{Index} 
class also provides an iterator over all keys, as well 
as a means of counting the number of rows with a given 
key.

The API is straightforward, but it is flexible, powerful and 
extensible. This simplicity ensures that 
programmer of all experience levels  
can take full advantage of the powerful data 
processing that wormtable provides. In the following subsections 
we illustrate the performance advantages of wormtable via some 
examples. These are not intended to be definitive benchmarks
and are simply concrete examples to demonstrate the type of improvements 
that can be expected by using wormtable over existing methods.

\subsection*{Scan performance}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable when performing calculations over 
a whole table, we converted a large 
publicly available VCF file (produced as part of the
Drosophila genetic reference panel~\cite{mr+12}) to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a workstation with an Intel Xeon processor, 12GB of RAM 
and a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).

@@

Full table scans are often uneccessary in wormtable. For many tasks,
creating an appropriate index allows us to seek directly to the 
rows of interest. To illustrate this, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m40s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

\subsection*{Seek performance}
One of the most serious problems with tabular data stored in text files
is that accessing a particular row efficiently without some auxiliary 
index. Without some information on where a particular row is located
in a file, there is nothing that can be done except to read the file 
line-by-line until the required row is found. Tabix~\cite{li11} solves
this problem by compressing a tab-delimited text file into blocks, and 
then storing an index mapping genome position to the location 
of the compressed block, and the position of the row within the block.
Tabix is specifically designed for range queries, allowing us to 
efficiently retrieve all of the rows within a given genomic range.

To compare the seek performance of wormtable with Tabix, we compressed
the VCF file mentioned in the previous subsection using \texttt{bgzip}
and indexed it using \texttt{tabix}. Since the VCF file was not 
strictly in order of genome position (in violation of the VCF format),
we first needed to sort the VCF file, which required 5m13s. Afterwards,
compression required 11m8s, and indexing with \texttt{tabix} required
1m12s. The size of the resulting compressed file was 2.9GB and 
the size of the index file was 106KB.
To duplicate the functionality of Tabix in wormtable, we simply need to 
create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
This required approximately
2 minutes and the resulting index consumed 115.4MB of space.

We compared the seek performance of Tabix and wormtable by generating 
a genomic location randomly and retrieving all rows within 1Kb of
this location. This 
was repeated $10^4$ times, and we measured both the elapsed and 
processor time. The Tabix Python module was used for the 
comparison, ensuring that no overheads associated with process 
forking were incurred. The same set of random locations were used 
for wormtable and Tabix, ensuring a fair comparison.

When using a cold cache (i.e., no pages of the files in question are 
present in the operating system's cache) the elapsed
time for Tabix was 103 seconds with a processor time of 65
seconds. For wormtable, the elapsed time was 181 seconds 
with a processor time of 5 seconds. Thus, the time required 
to perform this test is dominated by waiting for I/O in both 
cases. Since the wormtable file is considerably larger than the 
compressed file used by Tabix, more and larger seeks were required 
to bring the required pages into memory. However, once the 
pages were in memory, Tabix needed to do much more work to 
decompress then and make them usable, as shown by the 
difference in processor times.

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevent pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, compression is a double-edged sword: 
it does lead to much smaller files, which reduces the number of random seeks 
required.  On the other hand, though, decompression is expensive and must 
be repeated each time a block is accessed, unless some sophisticated
caching mechanism is devised.

The penalty for random seeks is exceedingly high in this 
example, however, since we have used a hard disk for storage.
Repeating the cold cache example on a laptop equipped with 
solid state storage, the elapsed time for Tabix was 76 seconds
with a CPU time of 72 seconds, whereas the elapsed time for 
wormtable was 31 seconds with a CPU time of 6.5 seconds. Thus, using 
solid state storage negates the advantage of smaller files, since 
seeks are much less expensive.  
 With the rapidly reducing cost and 
increasing capacity of solid state storage, it seems likely 
that it will soon be routinely used for all but the largest 
of bulk storage. Wormtable takes much better advantage of
modern hardware and large cache sizes, allowing the full bandwidth 
and random seek performance to be utilised. Compression based 
techniques will always be limited by the CPU time required to 
decompress data.

\section*{Discussion}
The problems of enabling efficient random access to rows and 
avoiding the large overhead of parsing text are well understood, 
and efforts to address them are proceeding in parallel for 
different file formats. 
BCF, for example, is the binary version of the VCF format discussed 
above, in which values within rows are stored in a packed binary 
format.  Similiarly, BigBed and 
BigWig~\cite{kzbhk10} are compressed binary versions of the BED 
and WIG file formats, which offer efficient random access along 
with the ability to operate over a network.

There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for many different file formats is surely 
unsustainable.

@@

The library supports efficient access to any data stored in 
wormtable format, and currently provides conversion utilitites 
for the VCF and GTF formats.
The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF and BED to wormtable format. Such tools are 
not difficult to develop, since all that is required is a parser
for the format in question written in Python.

Wormtable is currently limited to supporting Python only, and 
another important aspect of future development is to 
develop a C library along with bindings for other popular
languages such as Perl, R and PHP. Compression of data can
result in poor performance, but it is often necessary when 
volumes of data are very large. Thus, we plan on introducing 
optional compression of the data file in wormtable in 
a future release. Beyond these additions, it is difficult
predict the precise direction of future development since this depends 
on feedback from the community.  Wormtable is an open and 
collaborative project actively seeking feedback and 
contributors.

@@

We have concentrated on the VCF format here since it is a widely 
used and popular format. After converting the VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the columns of 
interest. For accessing regions of the genome efficiently
the user simply needs to create an index on the CHROM and POS columns
using \texttt{wtadmin add} command, and can then efficiently
seek to arbitrary locations. In the worst case, when using a 
slow hard disk this is somewhat slower than Tabix. In the best
case, when frequently accessing the same regions or using 
fast solid state storage, wormtable is many times faster than 
Tabix. Moreover, wormtable is much more general than Tabix, 
since it can index any column or combination of columns.
Similarly, wormtable is not limited to VCF, but can store
any form of fixed tabular data. We also provide a conversion 
tool for the GTF format, and several others are planned.

Wormtable is not intended to replace text files as the universal 
interchange format for biological data. 
It is intended to provide a persistent data structure 
that can be efficiently processed 
and searched. Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write straightforward code to process 
data efficiently. 
Different file formats can be handled consistently 
in wormtable, reducing the need for third party libraries 
to parse complex files and simplifying the code required 
to process data.
Together, these advantages of increased performance and reduced 
code complexity can substantially increase a researcher's 
productivity and ability to explore their data.

%%%%%%%%%%%%%%%%%%
\section*{Availability and requirements}
\paragraph{Project name:} Wormtable.
\paragraph{Project home page:}\url{https://pypi.python.org/pypi/wormtable}
\paragraph{Operating system(s):} POSIX.
\paragraph{Programming language:} C, Python.
\paragraph{Other requirements:} Berkeley DB 4.8 or later.
\paragraph{License:} GNU LGPLv3+.
\paragraph{Any restrictions to use by non-academics:} None.
